---
pagetitle: Expectations
output: 
  revealjs::revealjs_presentation:
    incremental: false
    theme: solarized
    self_contained: false
    # reveal_plugins: ["menu","notes","chalkboard"]
    # reveal_plugins: ["menu"]
    highlight: pygments
    center: true
    transition: none
    background_transition: none 
    reveal_options:
      # chalkboard:
      #   theme: whiteboard
      #   toggleNotesButton: true
      #   toggleChalkboardButton: true
      menu:
        numbers: true
      slideNumber: true
      previewLinks: false
    fig_caption: true
    css: mystyle.css
    
--- 

<section>

<h1>Expectations</h1>

Based on Stock and Watson, ch. 2

<br>

<h2>[Jesper Bagger](mailto:jesper.bagger@rhul.ac.uk)</h2>

<h3>EC2203 | Royal Holloway | 2020/21</h3>



</section>

# Expectations

## The expected value of a random variable

- The **expected value** of $Y$, denoted $\textrm{E}(Y)$, is the average value of $Y$ over many repeated, independent occurrences 

  <div class="box">
  $$\textrm{E}(Y) = \left\{ 
  \begin{array}{ll}
  \sum_y y f_Y(y) & \text{if } Y \text{ discrete} \\[0.5ex]
  \int y f_Y(y) dy & \text{if } Y \text{ continuous}
  \end{array}
  \right.$$
  </div>

- $\textrm{E}(Y)$ is the **mean** of $Y$, and is sometimes denoted $\mu_Y$

## The mean $\mu_Y$ measures distribution location

```{r echo=FALSE, error=FALSE, warning=FALSE}
# draw a plot of the N(0,1) PDF
curve(dnorm(x),
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-3.,4 ),
      ylab = "Density",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
curve(dnorm(x,mean = 1, sd = 1),
      col = "red",
      lwd = 3,
      add = T)
# add a legend
legend("topright",
       as.character(c("Mean = 0","Mean = 1")), 
       col = c("blue","red") , 
       lwd = 3,
      bty = "n",
       cex = 1.5)

```


## Variance of a random variable

- The **variance** of $Y$, denoted $\textrm{var}(Y)$, is the expected value of the squared deviation of $Y$ from its mean:

  <div class="box">
  $$\mathrm{var}(Y) = \textrm{E}[(Y-\mu_Y)^2]$$
  </div>

- The variance is often denoted $\sigma^2_Y$

## The variance $\sigma^2_Y$ measures distribution spread

```{r echo=FALSE, error=FALSE, warning=FALSE}
# draw a plot of the N(0,1) PDF
curve(dnorm(x),
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-6.5,6.5),
      ylab = "Density",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
curve(dnorm(x,mean = 0, sd = 2),
      col = "red",
      lwd = 3,
      add = T)
# add a legend
legend("topright",
       as.character(c("Var = 1","Var = 4")), 
       col = c("blue","red") , 
       lwd = 3,
      bty = "n",
       cex = 1.5)

```


## The standard deviation of a random variable

- The **standard deviation** is the square root of the variance

  <div class="box">
  $$\sigma_Y = \sqrt{\sigma^2_Y}$$
  </div>

- The standard deviation is measured in units of $Y$


## A standardized random variable 

- **Standardization** transforms a random variable $Y$ into a random variable with mean 0 and variance 1

- The standardized random variable computed from $Y$ is

  <div class="box">
  $$\frac{Y - \mu_Y}{\sigma_Y}$$
  </div>

- Standardized random variables are **unit-free**

## The Normal distribution

- Suppose $Y$ has a **Normal distribution** with mean $\mu_Y$ and variance $\sigma_Y^2$. We write

  $$Y \sim \mathcal{N}(\mu_Y,\sigma_Y^2)$$
- If we standardize $Y$, we obtain random variable $Z$ with a **standard Normal distribution**; we write

  $$Z = \frac{Y-\mu_Y}{\sigma_Y} \sim \mathcal{N}(0,1)$$

- The standard Normal PDF is denoted $\phi(z)$; the standard Normal CDF is $\Phi(z) = \Pr(Z \leq z) = \int_{-\infty}^z \phi(x) dx$


## The standard Normal PDF

```{r echo=FALSE, error=FALSE, warning=FALSE}
# draw a plot of the N(0,1) PDF
curve(dnorm(x),
      cex.axis = 1.5, 
      cex.lab = 1.5, 
      xlim = c(-6.5,6.5),
      ylab = "Density",
      col = "blue",
      lwd = 3,
      panel.first = grid(nx = NULL, 
                        ny = NULL, 
                        col = "lightgray", 
                        lty = "dotted", 
                        lwd = 2)) 
box(which = "plot", 
    lty = "solid", 
    lwd = 3)
curve(dnorm(x,mean = 1, sd = 2),
      col = "red",
      lwd = 3,
      add = T)
# add a legend
legend("topright",
       as.character(c("N(0,1)","N(1,4)")), 
       col = c("blue","red") , 
       lwd = 3,
       bty = "n",
       cex = 1.5)

```


# Summary

## Summary

- The expected value of a random variable is the probability or density weighted average of the random variable

- The mean of a random variable measures the location of its distribution 

- The standard deviation of a random variable measures the spread of its distribution

- A standardized random variable has unit mean and unit variance
